%@descr: A template for a paper, a report or a thesis - suitable for modifications
%@author: Maciej Komosiński

\documentclass{article} 
\usepackage[english]{babel} 
\usepackage[utf8]{inputenc} 
\usepackage[OT4]{fontenc}
\usepackage{graphicx,color} %include pdf's (and png's for raster graphics... avoid raster graphics!)
\usepackage{amsmath} %just for \eqref
\usepackage{url}
\usepackage{subcaption} % or \usepackage{subfigure}
\usepackage{float} % for [H] placement specifier
\usepackage{multicol}
\usepackage[pdftex,hyperfootnotes=false,pdfborder={0 0 0}]{hyperref} % after all packages; pdfborder not implemented the same way for every implementation because the specification is imprecise; under miktex you just don't see the frames


\input{_settings}

%\title{Lab report:\\Biologically-inspired algorithms and models}
%\author{}
%\date{}


\begin{document}
\input{_title}

\section{Problem description}\label{sec:problem-description}
The Quadratic Assignment Problem (QAP) is a combinatorial optimization problem focused on assigning a set of facilities to a set of locations
in a way that minimizes the total cost.
Each facility has an associated flow and each pair of facilities has an associated distance or cost.
The objective is to minimize the sum of the products of the flows and distances, considering the assignment of facilities to locations.

Applications of QAP span various fields including logistics, telecommunications, computer science, and facility layout planning.
In logistics, it's used for warehouse organization and facility allocation.
In telecommunications, it aids in network optimization.
In computer science, it's applied to circuit design and compiler optimizations.
QAP also finds relevance in operations research for optimal facility placement.

Interpretations of QAP can range from arranging physical facilities to minimize transportation costs to organizing data to reduce access time.
Its complexity is significant, classified as NP-hard, making it challenging to solve for large instances.

\subsection{Problem formulation}\label{subsec:problem-formulation}
The Quadratic Assignment Problem can be formulated as follows:
Given $n$ facilities and $n$ locations, where $n$ is the number of facilities and locations,
the goal is to find a permutation $\pi$ of the facilities that minimizes the total cost.
The total cost is calculated as the sum of the products of the flows and distances for all pairs of facilities.
The cost function is defined as:
\[ f(\pi) = \sum_{i=1}^{n} \sum_{j=1}^{n} a_{ij} \cdot b_{\pi(i)\pi(j)} \]
where $a_{ij}$ is the flow between facilities $i$ and $j$, $b_{\pi(i)\pi(j)}$ is the distance between the locations assigned to facilities $i$ and $j$,
and $\pi$ is a permutation of the facilities.

The goal is to find the permutation $\pi$ that minimizes the total cost $f(\pi)$.

\subsection{Instance description}\label{subsec:instance-description}
The instances used in this study are taken from the QAPLIB library, a collection of QAP instances used for benchmarking algorithms.
The instances used in this study are:
\begin{itemize}
    \item \textit{chr12a}: A 12-facility instance with known optimal solution.
    \item \textit{chr15a}: A 15-facility instance with known optimal solution.
    \item \textit{chr18a}: An 18-facility instance with known optimal solution.
    \item \textit{chr20a}: A 20-facility instance with known optimal solution.
    \item \textit{chr22a}: A 22-facility instance with known optimal solution.
    \item \textit{chr25a}: A 25-facility instance with known optimal solution.
    \item \textit{lipa20a}: A 20-facility instance with known optimal solution.
    \item \textit{tai12b}: A 12-facility instance with known optimal solution.
    \item \textit{tai100b}: A 100-facility instance with known optimal solution.
\end{itemize}

The selection of instances from the \textit{chr} group was based on their progressive difficulty levels,
allowing for a comprehensive evaluation spanning from relatively easier to more challenging problem instances.
Additionally, the inclusion of the \textit{lipa20a} instance, despite its similar size to the \textit{chr} instances,
was motivated by its distinct characteristics, providing a different set of challenges for algorithm evaluation.
While generally less demanding than the \textit{chr} instances, it offers valuable insights into algorithm performance across diverse problem landscapes.
Furthermore, the inclusion of the \textit{tai} instances was aimed at testing the algorithms across a wider spectrum of problem sizes.


\section{Implementation}\label{sec:implementation}
The Quadratic Assignment Problem was solved using five algorithms:
\begin{itemize}
    \item \textbf{Heuristic (H):} The Heuristic algorithm operates by analyzing the sums of rows in matrices $a$ and $b$, representing flows and distances.
        It iteratively identifies the facility with the highest sum in matrix $a$ and assigns it to the location with the lowest sum in matrix $b$.
        This process continues until a solution is reached.
    \item \textbf{Steepest (S):} A local search algorithm that iteratively explores the neighborhood of the current solution and selects the best neighbor.
        It also allows for some number of subsequent non-improving moves before terminating.
    \item \textbf{Greedy (G):} A local search algorithm that iteratively explores the neighborhood of the current solution and selects the first improving neighbor.
        It terminates when no improvement is possible.
    \item \textbf{Random Walk (RW):} An algorithm that in each step selects a random neighbor and track the best solution encountered.
    \item \textbf{Random Search (RS):} A random search algorithm that in each step generates random permutations of the facilities and selects the best solution encountered.
\end{itemize}

Implemenatation of these algorithms was done in Rust programming language.


\section{Description of the neighborhood operators}\label{sec:description-of-the-neighborhood-operators}
For the neighborhood operator ``Swap'' operator was used.
In this operator, two elements within a permutation representing the assignment of facilities to locations are selected,
and their positions are swapped.
This operation effectively generates a neighboring solution by altering the assignment of two facilities to different locations,
potentially leading to a decrease in the total cost.
The neighborhood size of the ``Swap'' operator corresponds to the number of possible swaps that can be performed within a given permutation.
For a permutation of length $n$, there are $\frac{n \times (n-1)}{2}$ possible swaps, as each pair of elements can be swapped,
and redundant swaps (swapping an element with itself or swapping two elements back to their original positions) are excluded.


\section{Algorithms Performance Comparison}\label{sec:algorithms-performance-comparison}
\input{algorithms-performance-comparison}

\section{Conclusions}\label{sec:conclusions}
The Quadratic Assignment Problem (QAP) poses a significant challenge in combinatorial optimization, finding applications across diverse fields.
This study evaluated the performance of five algorithms - Heuristic, Steepest, Greedy, Random Walk, and Random Search - on a range of QAP instances.
The results demonstrate that the Greedy and Steepest algorithms consistently outperform the Random Walk and Random Search algorithms in terms of solution quality.
Although the Heuristic algorithm yields suboptimal solutions,
its advantage lies in its speed, making it suitable for less challenging instances prioritizing efficiency.
Both the Greedy and Steepest algorithms emerge as top performers in solution quality.
However, the choice between them should consider specific problem requirements, as the difference in performance is not substantial.
Conversely, the Random Walk and Random Search algorithms deliver poor results, with solution quality deteriorating as instance difficulty increases.
Variations in the number of algorithmic steps and evaluated solutions are observed among algorithms,
with the Greedy algorithm conducting fewer evaluations compared to Steepest and Random Walk algorithms.
Furthermore, for the Greedy and Steepest algorithms, the quality of the solution is found to be independent of the initial solution quality.
The stability of average and lowest solution costs after a certain number of runs suggests the benefit of running the algorithm for at least 10 to 20 iterations.
Regarding the relationship between solution quality and distance from the optimal solution,
a negative correlation is observed for both instances tai12b and tai100b.
This indicates that solution quality tends to decrease as the distance from the optimal solution increases.
However, the lack of significance in the correlation implies that local minima are scattered throughout the decision space.

In summary, while the Greedy and Steepest algorithms exhibit superior performance,
careful consideration of algorithmic characteristics and problem requirements is essential for algorithm selection.
Further research could explore algorithmic enhancements and strategies for addressing challenges posed by local minima.

\section{Future Work}\label{sec:future-work}
\input{future-work}

\end{document}
